--- 
title: "Harvested Wood Products Carbon Model, Version R Documentation"
author: "| Jeremy Groom, Ph.D.\n| Groom Analytics, LLC \n| \n| Nadia Tase\n| California Department of Forestry and Fire Protection\n"
date: "`r paste('Date:',format(Sys.Date(), '%B %d, %Y'))`"
output: pdf_document
site: bookdown::bookdown_site
documentclass: book
classoption: openany
bibliography:
- book.bib
- packages.bib
description: "This HWP vR documentation was written in bookdown.  \nThe HTML output
  format for this document is bookdown::gitbook,\nset in the _output.yml file.\n"
link-citations: yes
github-repo: jeremygroom/HWP-vR-Documentation
header-includes:
- \usepackage[fontsize=12pt]{scrextend}
- \usepackage{titling}
- \pretitle{\begin{center}\vspace{5cm}\LARGE}
- \posttitle{\end{center}}
- \preauthor{\begin{center}\vspace{1.5cm}\Large\textit}
- \postauthor{\end{center}}
- \predate{\begin{center}\vspace{1cm}\Large\textit}
- \postdate{\end{center}\begin{center}\vspace{2cm}\includegraphics[width=3cm]{GroomAnalyticsE.png}\end{center}}

---

```{r libraries, echo = FALSE, warning=FALSE}
library(readr)
library(kableExtra)


```



# Summary       {#sum}

The Harvested Wood Products (HWP) model calculates cumulative HWP carbon stocks and emissions using the [Tier 3 Production Approach](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_04_Ch4_Forest_Land.pdf) carbon estimation guidelines developed by the [Intergovernmental Panel on Climate Change](https://www.ipcc.ch/).  It is a deterministic model which relies on optional and required user-supplied information.  The model estimation methods are described in @stockmann2012.  There have been several builds of the model, originally created by the U.S. Forest Service (USFS), on platforms including Microsoft Excel and C++. Through partnership with the California Department of Forestry and Fire Protection (CAL FIRE), Oregon Department of Forestry (ODF) and Groom Analytics LLC, this version of the tool (deemed HWP-C vR) builds from the previous USFS versions on the R [@R-base] platform, for use in their state-level harvested wood product carbon inventories.  Another USFS effort is currently underway to construct an interactive version of the model in Python.  This version (HWP-C vR) is not intended to compete with or contradict the original or ongoing USFS work, but rather complement the work by being available in a different platform and coding language. This version also includes new data visualizations.    

Here we describe the HWP-C vR build of the model based on data from the U.S. west coast states of California, Oregon, and Washington.  Users may explore California and Oregon data via the [R Shiny web application](https://groomanalyticsllc.shinyapps.io/HWP-C-vR/).  Users from any state or ownership may also upload their own data onto the app, run their data through the HWP model and associated Monte Carlo simulation, view and interact with figures, and download figures and tables.    

Alternatively, users may access the full Shiny app R code as well as a stand-alone version of the R code for HWP model implementation. The stand-alone version allows users the opportunity to step through the code more easily, save intermediate outputs (i.e., arrays) for hand-calculation of outcomes, and alter the code to meet their own data needs.  Both versions allow users to download or generate the final output tables necessary to recreate figures to their own specifications.   

This document describes the model and structure of the code, explains how to generate input files, and provides instruction on accessing the code. 



## Acknowledgements    {#sum-ack}
Many thanks to Nadia Tase (California Department of Forestry and Fire Protection - CAL FIRE) and Andrew Yost (Oregon Department of Forestry - ODF) for supporting the development of this work through feedback, discussion, and their own efforts to dive into the code and understand the model at its core.  Jimmy Kagan (Oregon State University's Institute for Natural Resources) and Amanda Reynolds (CAL FIRE) provided administrative support.  This work would not be possible without the original work completed by Nate Anderson (United States Forest Service - USFS), Keith Stockmann (USFS), Dan Loeffler (University of Montana), and others. The information and feedback they shared has been invaluable.  Feedback and support for an earlier iteration of the R-version of the model from Todd Morgan (University of Montana, Bureau of Business and Economic Research) has also been indespensible.  

This project was funded by the California Department of Forestry and Fire Protection Purchase Order 3540-0000277546.  Earlier work on the project was funded by the Oregon Department of Forestry and supported by the Institute for Natural Resources, Oregon State University.  

## Abbreviations      {#sum-abb}

The following description uses many acronyms for brevity.  Here is a list of the most common:

BF: Board Feet   
BBF: Billion Board Feet   
BLM: Bureau of Land Management   
DEC: Discard Energy Capture   
EEC: Emitted with Energy Capture   
EWOEC: Emitted without Energy Capture   
EUR: End-use Product Ratios   
HWP: Harvested Wood Products   
CCF: Hundred Cubic Feet   
IPCC: Intergovernmental Panel on Climate Change  
MBF: Thousand Board Feet), 
MT: Metric Tons   
MMT C: Million Metric Tons of Carbon   
PPR: Primary Product Ratios   
PIU; Products in Use   
SWDS: Solid Waste Disposal Sites   
Tg C: Teragrams of Carbon: Tg C   
Tg CO2e: Teragrams of Carbon Dioxide Equivalent   
TPO: Timber Product Output   
TPR: Timber Product Ratios   
USFS: United States Forest Service   


A note about units for carbon:  
Publications from the United States have frequently described carbon storage and emissions using the metric million metric tons carbon, or MMT C.  We are using the equivalent metric, teragrams of carbon (Tg C) in the HWP model and this manual to align with common scientific reporting units.  Additionally, though the tool provides estimates for carbon emissions in both units of carbon and carbon dioxide equivalents, these estimates are for the biogenic carbon content in the wood only, and do not include other carbon-based greenhouse gas emissions with different global warming potentials, such as methane.  
  




 
```{r include=FALSE}
# automatically create a bib database for R packages
#knitr::write_bib(c(
#  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lhs', 'shiny'#, 'renv'
#), 'packages.bib')
```
